# Basic R - plotting

## Basic embedded plot types

Looking at numbers is boring - people want to see pictures! Doing analyses without visualizations is like only listening to a movie.

There are a few plot types supported by base R that deal with (combinations of) vectors:  

- scatter (or line-) plot
- barplot
- histogram
- boxplot

We'll only look at the bare basics in this chapter because we are going to do it for real with package `ggplot2` in the next course. 


### Scatter and line plots

Meet `plot()` - the workhorse of R plotting.

```{r plot-hello-1, out.width='80%', fig.asp=.75, fig.align='center', fig.cap='Here is a nice figure!'}
time <- c(1, 2, 3, 4, 5, 6)
response <- c(0.09, 0.30, 0.41, 0.48, 0.72, 1.12)
plot(x = time, y = response)
```

The function plot is used here to generate a _scatter plot_. It may generae other types of figures, depending on its input as we'll see later.

#### Formula notation {-}

Instead of passing an `x =` and `y =` set of arguments, it is also possible to call the plot fuction with a **_formula notation_**:

```{r plot-hello-with-formula, out.width='80%', fig.asp=.75, fig.align='center'}
plot(response ~ time)
```

You can read `response ~ time` as _response as a function of time_. This is a nice, readable alternative in this case, but for many functions it is the only or preferred way to specify the relationship you want to investigate.

#### Plot decorations {-}

Plots should always have these decorations: 

- Axis labels indicating measurement type (quantity) and its units. E.g. '[Mg] (mq/ml)' or 'Heartrate (bpm)'.
- If multiple data series are plotted: a legend
- Either a title or a figure caption, depending on the context.

The first plots of this chapters were very bare (and a bit boring to look at): the plot has no axis labels (quantity and units) and no decoration whatsoever. By passing arguments to `plot()` you can modify or add many features of your plot. Basic decoration includes  

- adjusting markers (`pch = 19`, `col = "blue"`)
- adding connector lines (`type = "b"`) or removing points (`type = "l"`)
- adding axis labels and title (`xlab = "Time (hours)"`, `ylab = "Systemic response"`, `main = "Systemic response to agent X"`)
- adjusting axis limits (`xlim = c(0, 8)`)

This is not an exhaustive listing; these are listed in the last section of this chapter.

Here is a more complete plot using a variety of arguments.

```{r plot-hello-2, out.width='80%', fig.asp=.75, fig.align='center'}
plot(x = time, y = response, pch = 19, type = "b", xlim = c(0, 8),
     xlab = "Time (hours)", ylab = "Systemic response (a.u.)",
     main = "Systemic response to agent X", col = "blue")
```


#### Adjusting the plot symbol {-}

When you have many data points they will overlap. Using transparency with the `rgb(,, alpha=)` color definition and/or smaller plot symbols (`cex=`) solves this.

```{r plot-hello-3, out.width='80%', fig.asp=.75, fig.align='center'}
x <- rnorm(1000, 10, 2); y <- x + rnorm(1000, 0.5, 0.5)
plot(x, y, pch = 19, cex = 0.6,
     col = rgb(red = 0, green = 0, blue = 1, alpha = 0.2))
```


### Barplots

Barplots can be generated in several ways:

- By passing a factor to `plot()` - it will generate a barplot of level frequencies. This is a shorthand for `barplot(table(some_factor))`.
- By using `barplot()`. The advantage of this is that accepts some graphical parameters that are not relevant and accepted by `plot()`, such as `beside =`, `height = `, `width = ` and others (type `?barplot` to see all).

Here is an example:

```{r barplot-0, out.width='80%', fig.asp=.75, fig.align='center'}
persons <- as.factor(sample(c("male", "female"), size = 100, replace = T))
plot(persons)
```


#### `barplot()` with a vector {-}

The function `barplot()` can be called with a vector specifying the bar heights (frequencies), or a `table` object.

```{r barplot-1, out.width='80%', fig.asp=.75, fig.align='center'}
frequencies <- c(22, 54, 12, 29)
barplot(frequencies, names = c("one", "two", "three", "four"))
```

With a table object:

```{r barplot-2, out.width='80%', fig.asp=.75, fig.align='center'}
table(persons)
barplot(table(persons))
```


#### `barplot()` with a 2D table object {-}

Suppose you have this data:

```{r barplot-3,, out.width='80%', fig.asp=.75, fig.align='center'}
set.seed(1234) 
course <- rep(c("biology", "chemistry"), each = 10)
passed <- sample(c("Passed", "Failed"), size = 20, replace = T)
tbl <- table(passed, course) # the order matters!
tbl
```

The `set.seed(1234)` makes the _sampling_ reproducible, although that sounds really unlogical. Discussing **_pseudorandom_** sampling is not within the scope of this course however.

You can create a **_stacked bar chart_** like this.

```{r barplot-4, out.width='80%', fig.asp=.75, fig.align='center'}
barplot(tbl, 
        col = c("red", "darkblue"), 
        xlim = c(0, ncol(tbl) + 2), 
        legend = rownames(tbl))
```

The `xlim =` setting is a trick to get the legend beside the plot.

Using the `beside = TRUE` argument, you get the bars **_side by side_**:

```{r barplot-5, fig.height = 3.5, fig.width = 5}
barplot(tbl, 
        col=c("red", "darkblue"), 
        beside = TRUE, 
        xlim=c(0, ncol(tbl)*2 + 3), 
        legend = rownames(tbl))
```

Later, we'll see another data structure to feed to barplot: the matrix.

### Histograms

Histograms help you visualise the distribution of your data. 

```{r histogram-1, out.width='80%', fig.asp=.75, fig.align='center'}
male_weights <- c(rnorm(500, 80, 8)) ## create 500 random numbers around 80
hist(male_weights)
```

Using the `breaks` argument, you can adjust the bin width. Always explore this option when creating histograms!

```{r histogram-2, out.width='80%', fig.asp=.75, fig.align='center'}
par(mfrow = c(1, 2)) # make 2 plots to sit side by side
hist(male_weights, breaks = 5, col = "gold", main = "Male weights")
hist(male_weights, breaks = 25, col = "green", main = "Male weights")
```

If you want a more detailed 

### Density plot as alternative to `hist()`

When you want a bit more fine-grained view of the distribution you can use a plot of a density function; by adding a `polygon()` you can even have some nice shading under the line:

```{r density-plot-1, out.width='80%', fig.asp=.75, fig.align='center'}
plot(density(male_weights),
     main = "A density plot of male weights",
     col = "blue", lwd = 2)
polygon(density(male_weights), col="lightblue")
```


### Boxplots

This is the last of the basic plot types. A boxplot is a visual representation of the *5-number summary* of a numeric variable: minimum, maximum, median, first and third quartile. 

```{r}
persons <- rep(c("male", "female"), each = 100)
weights <- c(rnorm(100, 80, 6), rnorm(100, 75, 8))
#print 6-number summary (5-number + mean)
summary(weights[persons == "female"])
```

Boxplots tell the same story as histograms, but are less precise. however, they are excellent when you want to show a series of subsets split over some variable.

```{r boxplots-1, out.width='80%', fig.asp=.75, fig.align='center', results = "hold"}
par(mfrow = c(1, 2)) # make 2 plots to sit side by side
# create boxplots of weights depending on sex
boxplot(weights ~ persons, ylab = "weight")
boxplot(weights ~ persons, notch = TRUE, col = c("yellow", "magenta"))
```

Use `varwidth = TRUE` when you want to visualize the difference in group sizes.

### Adding more data and a legend 

When you have more than one data series to plot, add them using the function `points()`. You call this function _after_ you created the primary plot. Since there are multiple lines you will also need a legend.


```{r plot-legend-1, out.width='80%', fig.asp=.75, fig.align='center'}
response2 <- c(0.07, 0.10, 0.17, 0.28, 0.46, 0.61)
plot(x = time, y = response, pch = 19, type = "b",
     xlab = "Time (hours)", ylab = "Systemic response (a.u.)",
     main = "Systemic response to agent X", col = "blue")
points(x = time, y = response2, col = "red", pch = 19, type = "b")
legend(x = 1, y = 1.0, legend = c("one", "two"), col = c("blue", "red"), pch = 19)
```

The `legend()` function is _very_ versatile. Have a look at the docs!
In its most basic form you pass it a position (x and y), series names, colors and plot character.

### Helper lines and `lm()`

Adding helper lines can be used to aid your reader in grasping and interpreting your data story.
Use the function `abline()` for this.

There are four types of helper lines you might want to add to a figure:

- A horizontal line with `h =`: indicate some y-threshold
- A vertical line with `v =`: indicate x-threshold or mean or some other statistic
- A line with an intercept (`a =`) and a slope (`b =`): often used to indicate some expected response, or diagonal x = y
- A linear model, determined with the `lm()` function. The linear model object actually contains an intercept and a slope value which is taken by `abline()`.

In the following plot, these four basic helper lines are demonstrated:

```{r plot-helpers-1, out.width='80%', fig.asp=.75, fig.align='center'}
plot(x = time, y = response, pch = 19, type = "b",
     xlab = "Time (hours)", ylab = "Systemic response (a.u.)",
     main = "Systemic response to agent X", col = "blue")
#horizontal line
abline(h = 0.3, lty = 2, lwd = 2, col = "red")
#vertical line
abline(v = 4, lty = 3, lwd = 2, col = "darkgreen")
#line with slope
abline(a = -0.1, b = 0.3, lwd = 2, col = "purple")
#linear model 
abline(lm(response ~ time),  lwd = 2, col = "maroon")
```

## Graphical parameters to `plot()`

There are _many_ parameters that can be passed to the plotting functions. Here is a small sample and their possible values.

```{r}
series <- 1:20
plot(0, 0, xlim=c(1,20) , ylim=c(0.5, 7.5), col="white" , yaxt="n" , ylab="" , xlab="")

# the rainbow() function gives a nice palette across all colors
# or use hcl.colors() to specify another palette
# use  hcl.pals() to get an overview of available pallettes
colors = hcl.colors(20, alpha = 0.8, palette = 'viridis')

#pch
points(series, rep(1, 20), pch = 1:20, cex = 2)
#col
points(series, rep(2, 20), col = colors, pch = 16, cex = 3)
#cex
points(series, rep(3, 20), col = "black" , pch = 16, cex = series * 0.2)

#overlay to create new symbol
points(series, rep(4, 20), pch = series, cex = 2.5, col = "blue")
points(series, rep(4, 20), pch = series, cex = 1.5, col = colors)
 
#lty
for (i in 1:6) {
    points(c(-2, 0) + (i * 3), c(5, 5), col = "black", lty = i, type = "l", lwd = 3)
    text((i * 3) - 1, 5.25 , i)
}
#type and lwd
for (i in 1:4) {
    #type
    points(c(-4, -3, -2, -1) + (i * 5), rep(6, 4),
           col = "black", type = c("p","l","b","o")[i], lwd=2)
    text((i * 5) - 2.5, 6.4 , c("p","l","b","o")[i] )
    #lwd
    points(c(-4, -3, -2, -1) + (i * 5), rep(7, 4), col = "blue", type = "l", lwd = i)
    text((i * 5) - 2.5, 7.23, i)
}
#add axis
axis(side = 2, at = c(1, 2, 3, 4, 5, 6, 7),
    labels = c("pch" , "col" , "cex" , "combine", "lty", "type" , "lwd" ),
    tick = FALSE, col = "black", las = 1, cex.axis = 0.8)
```


