# Functions

## Dealing with NAs

Dealing with NA is a **_very big thing_**. When you work with external data there is always the possibility that some values will be missing.

You should be aware that it is not possble to test for `NA` values (Not Available) values in any other way; using `==` will simply return `NA`:

```{r}
x <- NA
x == NA
```

Other important functions for dealing with that are `na.omit()` and `complete.cases()`. Besides that, many functions in R have a (variant of) the `na.rm = ` argument. For nstance, when the `sum()` function encounters an NA in its input vector, it will always return `NA`:

```{r}
x <- c(1, 2, 3, NA)
sum(x)
sum(x, na.rm = TRUE)
```


## Descriptive statistics

R provides a wealth of descriptive statistics functions. The most important ones of them are listed below. The ones with an asterisk are described in more detail in following paragraphs.

| function        |  purpose             | 
|-----------------|----------------------| 
| `mean( )`       | mean                 | 
| `median( )`     | median               | 
| `min( )`        | minimum              | 
| `max( )`        | maximum              | 
| `range( )`      | min and max          | 
| `var( )`        | variance s^2         | 
| `sd( )`         | standard deviation s | 
| `summary( )`    | 6-number summary     | 
| `quantile( )` * | quantiles            | 
| `IQR( )`  *     | interquantile range  | 


#### The `quantile()` function {-}

This function gives the data values corresponding to the specified quantiles. The function defaults to the quantiles `0%  25%  50%  75% 100%`: these are the _quartiles_ of course.

```{r}
quantile(ChickWeight$weight)
quantile(ChickWeight$weight, probs = seq(0, 1, 0.2))
```

#### Interquantile range `IQR()` {-}

Function `IQR()` gives the range between the 25% and 75% quantiles.

```{r}
IQR(ChickWeight$weight)
## same as
quantile(ChickWeight$weight)[4] - quantile(ChickWeight$weight)[2]
## same as
diff(quantile(ChickWeight$weight, probs = c(0.25, 0.75)))
```

#### `boxplot()` is `summary()` visualized {-}

Boxplot is a graph of the 5-number summary, but `summary()` also gives the mean 

```{r box_demo, fig.height=2.5, fig.width=3.5}
summary(ChickWeight$weight)
boxplot(ChickWeight$weight)
```


## General purpose functions


#### Memory management {-}

When working with large datasets it may be usefull to free some memory once in a while (i.e. intermediate results). Use `ls()` to see what is in menory; use `rm()` to delete single or several items: `rm(genes)`, `rm(x, y, z)` and clear all by typing `rm(list = ls())`


#### File system operations {-}

Several functions exist for working with the file system:  

- `getwd()` returns the current working directory.
- `setwd(</path/to/folder>)` sets the current working directory.
- `dir()`, `dir(path)` lists the contents of the current directory, or of `path`.
- A `path` can be defined as `"E:\\emile\\datasets"` (Windows) or, on Linux/Mac using relative paths `"~/datasets"` or absolute paths `"/home/emile/datasets"`. 


#### Glueing character elements: `paste()` {-}

Use `paste()` to combine elements into a string

```{r paste-demo}
paste(1, 2, 3)
paste(1, 2, 3, sep="-")
paste(1:12, month.abb)
```

There is a variant, `paste0()` which uses no separator by default. 

#### A local namespace: `with()` {-}

When you have a piece of related code operating on a single dataset, use `with()` so you don't have to type its name all the time.

```{r with-demo, eval=FALSE}
with(airquality, {
  mdl <- lm(Solar.R ~ Temp)
  plot(Solar.R ~ Temp)
  abline(mdl)
})
```

Local variables such as `mdl` will not end up in the global environment. 

## Convert numeric vector to factor: `cut()`

Sometimes it is useful to work with a factor instead of a numeric vector. For instance, when working with a Body Mass Index (bmi) variable it may be nice to split this into a factor for some analyses.
The function `cut()` is used for this.
Suppose you have the following fictitious dataset

```{r, tidy = F}
## body mass index
bmi <- c(22, 32, 21, 37, 28, 34, 26, 29,
         41, 18, 22, 27, 32, 31, 26)
## year income * 1000 euros
income <- c(23, 14, 20, 13, 47, 15, 38, 29, 
            12, 25, 33, 24, 19, 42, 38)
my_data <- data.frame(bmi = bmi, income = income)
```

You can of course look at income as a function of bmi using a scatter plot:

```{r cut-demo-2, out.width='70%', fig.asp=.75, fig.align='center'}
with(my_data, plot(income ~ bmi))
```

But wouldn't it be nice to look at the bmi categories as defined by the WHO? To be able to do this, you need to split the numeric `bmi` variable into a factor using `cut()`.


```{r cut-demo-3, out.width='70%', fig.asp=.75, fig.align='center'}
my_data$bmi_class <- cut(bmi,
    breaks = c(0, 18.5, 25.0, 30.0, Inf), 
    right = F,
    labels = c("underweight", "normal", "overweight", "obese"),
    ordered_result = T)
with(my_data, boxplot(income ~ bmi_class))
```

The `breaks =` argument specifies the split positions; the `right = F` arguments specifies that the interval is _inclusive_ on the lower (left) boundary:

```{r}
x <- c(2, 5, 10)
cut(x, breaks = c(0, 2, 5, 10), right = F)
cut(x, breaks = c(0, 2, 5, 10), right = T)
```

An interval written as `(5,10]` means it is from -but excluding- 5 to -but including- 10. 
Note that in the first example the last value (10) becomes NA because 10 is exclusive in that interval specification.

## File I/O revisited

Whatever the contents of a file, you always need to address (some of) these questions:

- Are there comment lines at the top?
- Is there a header line with column names?
- What is the column separator?
- Are there quotes around character data?
- How are missing values encoded?
- How are numeric values encoded?
- Are there dates (a special challenge)
- What is the type in each column?
    - character / numeric / factor / date-time


#### Some `read.table()` arguments {-}

| arg              | specifies          | example           | 
|------------------|--------------------|-------------------| 
| **sep**          | field separator    | sep = ":"         | 
| **header**       | is there a header  | header = F        | 
| **dec**          | decimal format     | dec = ","         | 
| **comment.char** | comment line start | comment.char = "" | 
| **na.strings**   | NA value           | na.strings = "-"  | 
| **as.is**        | load as character  | as.is = c(1,4)    | 
| **stringsAsFactors** | load strings as factors | stringsAsFactors = F |


#### The data reading workflow {-}

Always apply this sequence of steps and repeat until you are satisfied with the result:

1. `read.table()` with arguments that seem OK
2. Check the result at least with `str()` and `head()` and verify that the columns have the correct data type.
    - Factors where numeric expected indicate missed "NA" values!
3. Adjust the read.table parameters
4. Rinse and repeat


#### Writing data to file {-}

To write a data frame, matrix or vector to file, use `write.table(myData, file="file.csv")`. Standard is a comma-separated file with both column- and row names, unless otherwise specified:  

- `col.names = F`
- `row.names = F`
- `sep = ";"`
- `sep = "\t" # tab-separated`

#### Saving R objects to file {-}

Use the `save()` function to write R objects to file for later use.
This is especially handy with intermediate results in long-running analysis workflows.

```{r eval = FALSE}
x <- stats::runif(20)
y <- list(a = 1, b = TRUE, c = "oops")
save(x, y, file = "xy.RData")
```


#### Writing a plot to file {-}

Use one of the functions `png()`, `jpeg()`, `tiff()`, or `bmp()` for these specific file types. They have widely differing properties, especially with repsect to file size.  
Use `width` and `height` to specify size. Default unit is pixels. Use other unit: `units = "mm"`

```{r eval = F}
png("/path/to/your/file.png",
    width = 700, height = 350, units = "mm")
plot(cars)
dev.off() # don't forget this one!
```

## Pattern matching 

>Pattern matching is the process of finding, locating, extracting and replacing patterns in character data that usually cannot be literally described.

For instance, it is easy enough to look for the word "Chimpansee" in a vector containing animal species names:

```{r}
animals = c("Chimpanzee", "Cow", "Camel")
animals == "Chimpanzee"
```

but what are you going to do if there are multiple variants of the word you are looking for? This?

```{r}
animals = c("Chimpanzee", "Chimp", "chimpanzee", "Camel")
animals == "Chimpanzee" | animals == "Chimp" | animals == "chimpanzee"
```

The solution here is not using literals, but to describe **_patterns_**.  

Look at the above example. How would you describe a pattern that would correctly identify all Chimpanzee occurrences?

Is you pattern something like this?

_A letter C in upper-or lower case followed by 'himp' followed by nothing or 'anzee'_

In programming we use **_regular expressions_** or **_RegEx_** to describe such a pattern in a formal consice way:

`[Cc]himp(anzee)?`

And to apply such a pattern in R, we use one of several functions dedicated for this task. Here is one, `grepl()`, which returns `TRUE` if the regex matched the vector element.

```{r}
grepl("[Cc]himp(anzee)?", animals)
```

#### Functions using regex {-}

There are  several functions dedicated to finding patters in character data. They differ in intent and output. Here is a listing.  

- **finding** Does an element contain a pattern (TRUE/FALSE)? `grepl(pattern, string)`
- **locating** Which elements contain a pattern (INDEX)? `grep(pattern, string)`
- **extracting** Get the content of matching elements `grep(pattern, string, value = TRUE)`
- **replace** Replace the first occurrence of the pattern `sub(pattern, replacement, string)`
- **replace all** Replace all occurrences of the pattern `gsub(pattern, replacement, string)`

Note that the `stringr` package from the tidyverse has many user-friendly functions in this field as well. Two of them will be dealt with in the exercises. 

### Regex syntax

A regular expression can be build out of any combination of  

- **character sequences** - Literal character sequences such as 'chimp'  
- **character classes** - A listing of possibilities for a single position. 
    - Between brackets: `[adgk]` means 'a' or 'd' or 'g' or 'k'. 
    - Use a hyphen to create a series: `[3-9]` means digits 3 through 9 and `[a-zA-Z]` means all alphabet characters. 
    - Negate using `^`. `[^adgk]` means anything _but_ a, d, g or k. 
    - A special case is the dot `.`: any character matches.  
    - Many special character classes exist (digits, whitespaces etc). They are discussed in a later paragraph.
- **alternatives** - Are defined by the pipe symbol `|`: "OR"   
- **quantifiers** - How many times the preceding block should occur. See next paragraph.  
- **anchors** - `^` means matching at the start of a string. `$` means at the end.

An excellent cheatsheet from the RStudio website is also included
<a href="figures/RegExCheatsheet.pdf" target="_blank">here</a>

### Quantifiers

Use quantifiers to specify how many times a character or series of characters should occur.

- **`{n}`**: exactly `n` times
- **`{n, }`**: at least `n` times
- **`{ ,n}`**: at most `n` times
- **`{n, m}`**: at least `n` and at most `m` times. 
- **`*`**: 0 or more times; same as `{0, }`
- **`+`**: 1 or more times; same as `{1, }`
- **`?`**: 0 or 1 time; same as `{0, 1}`

### Some examples

#### Restriction enzymes {-}

This is the recognition sequence for the _HincII_ restriction endonuclease:

```
5'-GTYRAC-3'
3'-CARYTG-5'
```

Before reading on: how would you define a regular expression that is precisely describes this recognition sequence?

Molecular biology seuence ambiguity codes can be found
<a href="data/ambiguity-codes.html" target="_blank">here</a>

```{r}
HincII_rs <- "GT[CT][AG]AC"
sequences <- c("GTCAAC",
               "GTCGAC",
               "GTTGAC",
               "aGTTAACa",
               "GTGCAC")
grep(pattern = HincII_rs, x = sequences, value = TRUE)
```

#### Dutch dates {-}

Here are some Dutch dates, in different accepted formats. The last two are not a correct notation.
Create a RegEx that will determine whether an element contains a Dutch date string.

```{r}
dates <- c("15/2/2019", "15-2-2019", "15-02-2019", "015/2/20191", "15/2/20191")
dateRegex <- "[0-9]{2}[/-][0-9]{1,2}[/-][0-9]{4}"
grep(pattern = dateRegex, x = dates, value = TRUE)
```

Why were the last two matched?  
Because the pattern _is there_, albeit embedded in a longer string.  
We have to **_anchor_** the pattern to be more specific. 

### Anchoring

Using anchoring, you can make sure the string is not longer than you explicitly state:

```{r}
dates <- c("15/2/2019", "15-2-2019", "15-02-2019", "015/2/20191", "15/2/20191")
dateRegex <- "^[0-9]{2}[/-][0-9]{1,2}[/-][0-9]{4}$"
grep(pattern = dateRegex, x = dates, value = TRUE)
```

Now the date matching is correct.

### Metacharacters: Special character classes

Since patterns such as `[0-9]` accur so frequently, they have dedicated character classes such as `[[:digit:]]`. The most important other ones are 

- **digits** `[[:digit:]]` or `\\d`: equivalent to `[0-9]`
- **alphabet characters** `[[:alpha:]]`: equivalent to `[a-zA-Z]`
- **lowercase characters** `[[:lower:]]`: equivalent to `[a-z]`
- **uppercase characters** `[[:upper:]]`: equivalent to `[A-Z]`
- **whitespace characters** `[[:space:]]` or `\\s`: Space, tab, vertical tab, newline, form feed, carriage return
- **punctuation characters** `[[:punct:]]`: One of !"#$%&’()*+,-./:;<=>?@[]^_`{|}~

(have a look at the cheat sheet for all) 

Here is the same example, this time using these predefined character classes

```{r}
dates <- c("15/2/2019", "15-2-2019", "15-02-2019", "15022019", "15/2/20191")
dateRegex <- "[[:digit:]]{2}[/-]\\d{1,2}[/-]\\d{4}"
grep(pattern = dateRegex, x = dates, value = TRUE)
```

#### Postal codes {-}

Here are some Dutch zip (postal) codes, in different accepted formats. The last two are not a correct notation.
Can you create a RegEx that will determine whether an element contains a Dutch zip code?

```{r}
zips <- c("1234 AA", "2345-BB", "3456CC", "4567 dd", "56789aa", "6789a_")
zips
```


#### Prosite patterns {-}

<a href="https://prosite.expasy.org/" target="_blank">Prosite</a> is a database of amino acid sequence motifs. One of them is the Histidine Triad profile (PDOC00694). 

```
[NQAR]-x(4)-[GSAVY]-x-[QFLPA]-x-[LIVMY]-x-[HWYRQ]-  
[LIVMFYST]-H-[LIVMFT]-H-[LIVMF]-[LIVMFPT]-[PSGAWN]
```

- Write this down as a RegEx
- Was that efficient? Using the `gsub()` function, can you convert it in a RegEx using code? It may take several iterations. Was that efficient?  
- Next, use an appropriate function to find if, and where, this pattern is located within the sequences in file `data/hit_proteins.txt` (<a href="data/hit_proteins.txt" target="_blank">here</a>)

Amino Acid codes and Prosite pattern encoding can be found
<a href="data/ambiguity-codes.html" target="_blank">here</a>


### Locating and Extracting 

This is outside the scope of this first acquaintance. The `stringr` package from the tidyverse is much better at this than the base R functions. One of the exercises introduces this topic for the eager ones among you.

